Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-25 article/article.Rmd
Commit 2024-06-25 21:11

We studied the enhancement of the global horizontal irradiance induced by clouds over
Thessaloniki for the period 1993 -- 2023 using data recorded every one minute.  We
identified the cloud enhancement events (CE) by creating an appropriate clear sky
irradiance reference with the use of a radiative transfer model and aerosol optical
depth data from a collocated Cimel sun photometer and a Brewer spectrophotometer. We
found a trend in CE events of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
and a trend in the CE events irradiation of 
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June.  An analysis of the total
duration of CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour.  Finally we have detected enhancements above
the total solar irradiance at the top of the atmosphere of up to
$400\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$200\,\text{W}/\text{m}^{2}$.

CE: cloud enhancement cases one minute

@@

$\text{GHI}_\text{CSm}$: modelled clear sky one minute global horizontal irradiance

@@

<!-- our work -->
In this study, we evaluate the effects CE on GHI by investigating their frequency of
occurrence, intensity, and duration in a thirty-year period of GHI observations at
Thessaloniki, Greece.  We used modeled clear sky irradiance as a baseline to identify
cloud enhancements and we determined long-term trends of the above mentioned metrics,
their climatology and some general characteristics.  To our knowledge there are no
other studies that provide trends from such  long dataset.

@@

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute GHI from the $\text{GHI}_\text{ref}$ corresponding to
cloud-free atmosphere.
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023], and
the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.  We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the CE events. However, by inspecting the daily plots of
irradiance it became evident that changes in atmospheric conditions introduced
numerous false positive and false negative results.  The main reason for these
discrepancies is the variability of the effects of aerosols and water vapor which
were not taken into account in the two simple methods.  To produce a more
representative reference  we included the effects of these factors through variables
using a radiative transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}..

@@

<!-- Libradtran options -->
We approximated the expected clear sky $\text{GHI}_\text{ref}$ with the radiative
transfer model uvspec, part of libRadtran [@Emde2016], similarly to the approach used
by @Vamvakas2020.  In uvspec we used the solar spectrum of @Kurucz1994 in the range
$280$ to $2500\,\text{nm}$, the radiative trasfer solver "disort" in
"pseudospherical" geometry and the "LOWTRAN" gas parameterization.  The model was run
for a range of variables in order to create a look up table (LUT) for the estimation
of the cloud-free reference irradiance for each individual observation of our
dataset. In this context, the model was run for SZAs in the range $10$ -- $90^\circ$
with a step of $0.2^\circ$ and for the atmospheric profiles of the Air Force
Geophysics Laboratory [@Anderson1986] midlatitude summer and midlatitude winter,
representative of the warm and cold seasons.

Main factors responsible for the attenuation of the broadband downward solar
radiation under cloud free atmospheres are aerosols and water vapor.  At
Thessaloniki, such measurements are available since 2003 from the Cimel sun
photometer that belongs to the Aerosol Robotic Network (AERONET) [@Giles2019;
@Buis1998].  From the observations in the period 2003 -- 2023 we calculated the
monthly climatological means and standard deviation ($\sigma$) for the aerosol
optical depth (AOD) at $500\,\text{nm}$ and the equivalent height of the water column
(WC).  The monthly climatological values of AOD and water column, as well as
combinations with additional offsets of $\pm1\sigma$ and $\pm2\sigma$ , were used as
inputs to the RTM in the construction of the LUT.

For each measurement of the dataset, we calculated from the LUT a
$\text{GHI}_\text{ref}$ value for the respective season and SZA (by linear
interpolation), and for the climatological values of AOD and WC of the respective
month. The same procedure was followed for the estimation of the
$\text{GHI}_\text{ref}$ for all combinations of the AOD and WC with the above
mentioned standard deviation offsets.  Finally, each $\text{GHI}_\text{ref}$ value
was adjusted to the actual Sun-Earth distance derived by the Astropy software library
[@AstropyCollaboration2022].

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method is capable in detecting all major CE events.  Where some CE events with
very low OIR may be not detected, these are few with small over-irradiance and it is
unlikely that will effect significantly our results.

<!-- extreme cases criterion -->
A sub-category  of the CE events that is often discussed in the relevant literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere. In this case the threshold $E$ is given by:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: E > \cos(\theta) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for
the actual Sun -- Earth distance $r$ and $r_\text{m}$ is the mean Sun -- Earth
distance of $1.496\times10^8\,\text{km}$.

@@

Following the application of the above discussed methodology to the entire dataset 
(`r round(ST_total$wattGLB.N / 10^6,1)` million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The strongest CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of
the CE events as well as of the corresponding OIR and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OIR and the yearly
excess irradiation  for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OIR is negligible 
($`r round(yearlyCE[var == "GLB_diff.mean", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OIR for the entire period is
$`r round(mean(ST_E_yearly$GLB_diff.mean), 1)`\,\text{W}/\text{m}^2 
\pm `r round(sd(ST_E_yearly$GLB_diff.mean), 1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

<!-- $`r round(yearlyCE[var == "GLB_diff.mean", slope], 2)`\pm -->
<!-- `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 2)`\,W/m^2/y$ -->
<!-- (Figure\nobreakspace{}\ref{fig:P-energy-mean}), -->
<!-- but this value alone is not very useful due the intrinsic high variability of this -->
<!-- metric. -->

<!-- ```{r CEmeanDaily2, echo=FALSE, fig.cap="Trend of the daily excess energy due to CE"} -->
<!-- knitr::include_graphics("../images/daily-12.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Treds of the mean over irradiance per CE"} -->
<!-- knitr::include_graphics("../images/energy-3.png") -->
<!-- ``` -->
<!-- ```{r P-energy-mean, echo=FALSE, fig.cap="Trends of the mean OIR per CE."} -->
<!-- knitr::include_graphics("../images/P-energy-3.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Trend of number of CE cases"} -->
<!-- knitr::include_graphics("../images/energy-2.png") -->
<!-- ``` -->
<!-- ```{r P-energy-N, echo=FALSE, fig.cap="Trend of yearly cloud enhancement occurrences."} -->
<!-- knitr::include_graphics("../images/P-energy-2.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Trend of the yearly excess energy due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/energy-1.png") -->
<!-- ``` -->
<!-- ```{r P-energy-sum, echo=FALSE, fig.cap="Trend of the yearly excess energy due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/P-energy-1.png") -->
<!-- ``` -->
<!-- ```{r P-energy-median, echo=FALSE, fig.cap="Trend of the yearly median over irradiance due to CE over irradiance"} -->
<!-- knitr::include_graphics("../images/P_energy-8.png") -->
<!-- ``` -->

@@

<!--
```{r echo=FALSE, fig.cap="Distribution of the number of CE cases"}
knitr::include_graphics("../images/climCEmonth-1.png")
```
--->

<!-- ```{r echo=FALSE, fig.cap="Distribution of CE above reference"} -->
<!-- knitr::include_graphics("../images/relative_distributions-2.png") -->
<!-- ``` -->

<!--
```{r echo=FALSE, fig.cap="Distribution of CE above reference"}
knitr::include_graphics("../images/relative_distributions-4.png")
```
-->

@@

By creating a clear sky approximation of the GHI, which represents the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
After analyzing the CE cases, we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.

The magnitude of the ECE events does not exceed the values reported for sites with
more favourable conditions for the phenomenon. The climatological characteristic of
the ECE events, showed that the most active months are spread in half of the year
(March -- June, September and October). We found that CE conditions, can have a
duration of more than an hour in rare cases, with the bulk of the cases having a
duration under 5 minutes.  Some of the characteristics of CE and ECE events we
analysed have strong similarities with results by @Vamvakas2020 for the city of
Patras, south of Thessaloniki, with analog differences on the intensity of the
solar radiation.

An interpretation of the CE trends shows that the interaction of GHI with the clouds,
through this 30 year period, is a dynamic phenomenon that needs further
investigation.  Although, to approach it, we need more long-term observations of AOD
and clouds characteristics.

>>>>

We studied the enhancement of the global horizontal irradiance induced by clouds over
Thessaloniki for the period 1993 -- 2023 using data recorded every one minute.  We
identified the cloud enhancement events (CE) by creating an appropriate clear sky
irradiance reference with the use of a radiative transfer model and aerosol optical
depth data from a collocated Cimel sun photometer and a Brewer spectrophotometer. We
found a trend in CE events of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
and a trend in the CE events irradiation of 
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June.  An analysis of the total
duration of CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour and up to 140 minutes.  Finally, we have detected enhancements above
the total solar irradiance at the top of the atmosphere of up to
$400\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$200\,\text{W}/\text{m}^{2}$. The most active season of these extreme events is
spring until early summer and autumn.

<!--
CE: cloud enhancement cases one minute

@@

$\text{GHI}_\text{CSm}$: modelled clear sky one minute global horizontal irradiance 
-->

@@

<!-- our work -->
In this study, we evaluate the effects CE on GHI by investigating their frequency of
occurrence, intensity, and duration in a thirty-year period of GHI observations at
Thessaloniki, Greece.  We used modeled clear sky irradiance as a baseline to identify
cloud enhancements, and we determined long-term trends of the above-mentioned metrics,
their climatology and some general characteristics.  To our knowledge there are no
other studies that provide trends from such long dataset.

@@

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute GHI from the $\text{GHI}_\text{ref}$ corresponding to
cloud-free atmosphere.
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023], and
the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.  We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the CE events. However, by inspecting the daily plots of
irradiance it became evident that changes in atmospheric conditions introduced
numerous false positive and false negative results.  The main reason for these
discrepancies is the variability of the effects of aerosols and water vapor which
were not taken into account in the two simple methods.  To produce a more
representative reference we included the effects of these factors through variables
using a radiative transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}.

@@

<!-- Libradtran options -->
We approximated the expected clear sky $\text{GHI}_\text{ref}$ with the radiative
transfer model uvspec, part of libRadtran [@Emde2016], similarly to the approach used
by @Vamvakas2020.  In uvspec we used the solar spectrum of @Kurucz1994 in the range
$280$ to $2500\,\text{nm}$, the radiative transfer solver "disort" in
"pseudospherical" geometry and the "LOWTRAN" gas parameterization.  The model was run
for a range of variables in order to create a look-up table (LUT) for the estimation
of the cloud-free reference irradiance for each individual observation of our
dataset. In this context, the model was run for SZAs in the range $10$ -- $90^\circ$
with a step of $0.2^\circ$ and for the atmospheric profiles of the Air Force
Geophysics Laboratory [@Anderson1986] midlatitude summer and midlatitude winter,
representative of the warm and cold seasons.

Main factors responsible for the attenuation of the broadband downward solar
radiation under cloud free atmospheres are aerosols and water vapor.  At
Thessaloniki, such measurements are available since 2003 from the Cimel sun
photometer that belongs to the Aerosol Robotic Network (AERONET) [@Giles2019;
@Buis1998].  From the observations in the period 2003 -- 2023 we calculated the
monthly climatological means and standard deviation ($\sigma$) for the aerosol
optical depth (AOD) at $500\,\text{nm}$ and the equivalent height of the water column
(WC).  The monthly climatological values of AOD and water column, as well as
combinations with additional offsets of $\pm1\sigma$ and $\pm2\sigma$, were used as
inputs to the RTM in the construction of the LUT.

For each measurement of the dataset, we calculated from the LUT a
$\text{GHI}_\text{ref}$ value for the respective season and SZA (by linear
interpolation), and for the climatological values of AOD and WC of the respective
month. The same procedure was followed for the estimation of the
$\text{GHI}_\text{ref}$ for all combinations of the AOD and WC with the
above-mentioned standard deviation offsets.  Finally, each $\text{GHI}_\text{ref}$
value was adjusted to the actual Sun-Earth distance derived by the Astropy software
library [@AstropyCollaboration2022].

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method is capable in detecting all major CE events.  Where some CE events with
very low OIR may be not detected, these are few with small over-irradiance, and it is
unlikely that will affect significantly our results.

<!-- extreme caces criterion -->
A sub-category of the CE events that is often discussed in the relevant literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere. In this case the threshold $E$ is given by:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: E > \cos(\theta) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for
the actual Sun -- Earth distance $r$ and $r_\text{m}$ is the mean Sun -- Earth
distance of $1.496\times10^8\,\text{km}$.

@@

Following the application of the above discussed methodology to the entire dataset 
($`r round(ST_total$wattGLB.N / 10^6,1)`$ million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The strongest CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of
the CE events as well as of the corresponding OIR and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OIR and the yearly
excess irradiation for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OIR is negligible 
($`r round(yearlyCE[var == "GLB_diff.mean", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OIR for the entire period is
$`r round(mean(ST_E_yearly$GLB_diff.mean), 1)`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

@@

@@

By creating a clear sky approximation of the GHI, which represents the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
After analyzing the CE cases, we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.
We found that continuous CE events can last up to $140$ minutes, and $80\,\\%$ of
them bellow $5$ minutes.

The magnitude of the ECE events does not exceed the values reported for sites with
more favourable conditions for the phenomenon. We have observed cases exceeding the
TSI by $400\,\text{W}/\text{m}^{2}$ with $75\,\%$ of the cases under
$200\,\text{W}/\text{m}^{2}$. The climatological characteristic of the ECE events,
showed that the most active months are spread in half of the year (March -- June,
September and October). We found that CE conditions, can have a duration of more than
an hour in rare cases, with the bulk of the cases having a duration under 5 minutes.
Some of the characteristics of CE and ECE events we analysed have strong similarities
with results by @Vamvakas2020 for the city of Patras, south of Thessaloniki, with
analog differences on the intensity of the solar radiation.

An interpretation of the CE trends shows that the interaction of GHI with the clouds,
through this 30-year period, is a dynamic phenomenon that needs further
investigation.  Although, to approach it, we need more long-term observations of AOD
and clouds characteristics.
